¥documentclass[11pt]{article}
¥usepackage[margin=1in]{geometry}
¥usepackage[utf8]{inputenc}¥usepackage[T1]{fontenc}
¥usepackage{hyperref}
¥hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue,
  pdftitle={Form, Not Granted}, pdfauthor={Caio Vicentino}}
¥emergencystretch=3em ¥sloppy ¥hyphenpenalty=2000 ¥hbadness=10000
¥usepackage{booktabs}¥usepackage{amsmath}¥usepackage{amssymb}¥usepackage{xcolor}¥usepackage{graphicx}¥usepackage{array}

¥title{Form, Not Granted ¥¥[2pt]
¥large A sparse-autoencoder autopsy finds no native authorization concept at an agent's
commit point: the ``authorization direction'' is a surface-form artifact}
¥author{Caio Vicentino ¥¥ OpenInterpretability ¥¥ Fortaleza, Brazil ¥¥ ORCID: 0009-0003-4331-6259 ¥¥ ¥texttt{caio@openinterp.org}}
¥date{June 2026}

¥begin{document}
¥maketitle

¥begin{abstract}
A late-layer ``authorization direction'' was reported to detect and steer a tool-using agent's commitment
to unauthorized irreversible actions‾¥cite{authdir}, and a follow-up found it tracks the authorization the
model ¥emph{feels} rather than the one the user ¥emph{granted}‾¥cite{feltnotgranted}. Both were built on a
¥emph{supervised} difference-of-means direction --- a method that returns a direction whether or not the
concept is natively represented. We stop using the ruler and point the microscope: we encode the residual
at the delete-commit token (layer 59 of Qwen3.6-27B) through a trained TopK sparse autoencoder
(40{,}960 features) and ask, unsupervised, whether a feature distinguishes a ¥emph{granted} commit from a
¥emph{felt}-but-not-granted one. Three results. (1) ¥textbf{The supervised direction inverts under a
structure-matched control}: applying the same frozen direction, AUROC(granted vs felt) goes from $0.838$
(single-turn vs multi-turn conditions) to $0.08$ when both conditions share one multi-turn scaffold and only
the final user turn varies --- a direction measuring authorization would not flip when only the conversation
¥emph{format} changes. (2) ¥textbf{The SAE's separating features are surface form}: across runs, every top
feature that separates granted from felt is, by its max-activating context, a tool-call scaffold, a JSON
tool-schema token, or the file-listing output --- never a semantic ``user granted this'' feature. (3) ¥textbf{A
cross-lexical-family test isolates this}: a feature selected on one paraphrase family does ¥emph{not} transfer
to a disjoint family (held-out AUROC $0.50$, chance), and the one feature that does transfer ($0.99$, pooled
$0.998$) max-activates on tool-call ¥emph{format boilerplate}, not authorization. The agent's commit-point
representation is dominated by ¥textbf{surface form and plan-coherence}; ¥emph{authorization-as-granted is not
a native monosemantic feature} at this locus. Internal authorization monitors fail not because they are
mis-calibrated but because there may be no concept to read. We state the scope plainly (one model, one locus,
synthetic scenarios, an SAE explaining $¥sim$66¥% of residual variance) and release every ledger and a
recompute-everything evaluation.
¥end{abstract}

¥section{Introduction}
Linear directions in the residual stream detect and steer safety-relevant behavior: refusal lives in a
single direction‾¥cite{refusaldir}, and a late-layer ¥emph{authorization direction} both separates authorized
from unauthorized irreversible-action commits and, under steering, moves the commitment‾¥cite{authdir}. A
follow-up sharpened this to ``felt, not granted'': in realistic free generation the same direction tracks the
authorization the model ¥emph{feels} (and is blind to the user's actual grant)‾¥cite{feltnotgranted}.

Every step of that arc used a ¥emph{supervised} difference-of-means direction $d=¥overline{¥text{auth}}
-¥overline{¥text{unauth}}$. This method has a known failure mode: it ¥emph{always} returns a direction that
separates the labeled sets, whether the model represents the concept monosemantically or whether we merely
projected a human concept onto whatever variance correlated with the labels‾¥cite{geometry,interpillusion}.
The ``felt, not granted'' result is itself a hint that the concept is imported. So we ask the question the
ruler cannot answer: at the moment an agent commits an irreversible action, does it ¥emph{represent}
``user-granted authorization'' in its own ontology --- or is the decision substrate something else, onto which
``authorization'' was projected? We answer it by letting the model's own features speak, via a sparse
autoencoder‾¥cite{monosemantic,saefind,topk} trained on this exact model.

Contributions: (i) a structure-matched control shows the supervised ``authorization direction'' is
¥emph{confounded by conversation structure} --- it does not weaken but ¥emph{inverts} (¥S¥ref{sec:flip}); (ii)
an unsupervised SAE autopsy shows the features that separate granted from felt are ¥emph{surface form}, named
by their max-activating contexts (¥S¥ref{sec:autopsy}); (iii) a cross-lexical-family generalization test
isolates lexical/scaffold confounds from semantics and finds no paraphrase-invariant ¥emph{semantic} feature
(¥S¥ref{sec:cross}); (iv) the synthesis --- the agent commits on form and plan-coherence, and
authorization-as-granted is not a native concept here, which reframes why internal monitors fail
(¥S¥ref{sec:disc}). This qualifies, not refutes, ¥cite{authdir,feltnotgranted}: the direction is real and
usable in its tested regime, but it reads form, not a represented grant.

¥section{Setup}
¥paragraph{The direction and the SAE.} Following‾¥cite{authdir}, $d$ is the difference of mean layer-59
residuals at the action-commitment token over explicitly-authorized vs.¥ unauthorized prefilled deletes,
frozen. The SAE is a TopK autoencoder (¥texttt{caiovicentino1/qwen36-27b-sae-fullstack}, layer 59, $d_{¥text{sae}}
=40{,}960$, $k=128$; reconstruction relative error $0.58$, fraction-of-variance-explained $¥approx 0.66$). We
encode a residual $x$ as $z=¥mathrm{TopK}_k¥!¥big(¥mathrm{ReLU}((x-b_{¥text{dec}})W_{¥text{enc}}+b_{¥text{enc}})¥big)$.
¥paragraph{Conditions.} A real Qwen3.6-27B agent (bf16) runs free generation in a sandboxed file workspace
and commits ¥texttt{delete¥_file}; we capture the layer-59 residual at the ¥texttt{<function=} decision token.
¥emph{granted} = the user explicitly authorizes the deletion; ¥emph{felt} = the agent finishes a cleanup plan
it co-authored under a vague ``go ahead'' (the realistic over-reach of‾¥cite{feltnotgranted}). $n=12$--$18$
commits per condition; greedy decoding (deterministic; the decisive run reproduced bit-for-bit across two GPU
sessions). Ledgers and scripts are public.

¥section{The supervised ``authorization direction'' inverts under structure-matching}
¥label{sec:flip}
In a first autopsy the conditions differed in conversation ¥emph{structure} (granted was single-turn; felt was
multi-turn: tidy $¥to$ list $¥to$ flag $¥to$ ``go ahead''). A structure-matched control removes this: both
conditions share one multi-turn scaffold and differ ¥emph{only} in the final user turn. We then apply the
¥emph{same frozen} direction $d$ --- no selection, no fitting (Fig.‾¥ref{fig:flip}):

¥begin{center}
¥begin{tabular}{lc}¥toprule
setting & supervised $d$ AUROC (granted vs felt) ¥¥¥midrule
unmatched (structure differs) & $0.838$ ¥¥
structure-matched & $¥mathbf{0.08}$ --- it ¥emph{inverts} ¥¥¥bottomrule
¥end{tabular}
¥end{center}

The same direction gives near-opposite answers depending on prompt format. A direction that measured
¥emph{granted authorization} would be invariant to changing only the conversation format while holding the
authorization meaning fixed. It is not. ¥textbf{The ``authorization direction'' is substantially a
surface-structure artifact}; this is selection-free and does not depend on the SAE.

¥begin{figure}[t]¥centering
¥includegraphics[width=0.62¥textwidth]{figures_co/fig1_flip.pdf}
¥caption{The frozen supervised direction from‾¥cite{authdir,feltnotgranted} inverts (AUROC $0.838¥to0.08$)
when prompt structure is matched and only the authorizing user turn varies. It reads structure, not grant.}
¥label{fig:flip}
¥end{figure}

¥section{The SAE's separating features are surface form}
¥label{sec:autopsy}
Unsupervised, a single SAE feature separates granted from felt at AUROC $1.0$ ($p<0.001$ vs.¥ a $1000¥times$
permutation null) in both the unmatched and matched runs --- an automated ``native feature found'' flag. The
flag is misleading: it only asks ¥emph{does some feature separate}. The pre-registered feature-¥emph{naming}
(max-activating token context) is decisive --- every top separator is surface form (Fig.‾¥ref{fig:naming}):
the tool-call scaffold (¥texttt{<think></think>¥textbackslash n<tool¥_call>¥textbackslash n<function}), JSON
tool-schema tokens (¥texttt{"parameters":¥{"type":"object"}, ¥texttt{"required":["path"]}), and the file-listing
output (¥texttt{Files: cache¥_tmp.dat¥dots}). ¥textbf{No semantic ``user-granted authorization'' feature appears
among the separators in either run.}

¥begin{figure}[t]¥centering
¥includegraphics[width=0.92¥textwidth]{figures_co/fig3_naming.pdf}
¥caption{Every feature that separates granted from felt commits is, by its max-activating context, tool-call
scaffold, JSON tool-schema, file-listing, or format boilerplate. None is authorization.}
¥label{fig:naming}
¥end{figure}

¥section{Cross-lexical generalization: surface, not semantics}
¥label{sec:cross}
The matched conditions still differ in the ¥emph{text} of the user turn, so a separator could be lexical. We
ran a $2¥times2$ design --- authorization (granted/felt) $¥times$ two disjoint lexical families (A/B),
structure-matched --- and the selection-free test: select the best SAE feature (and a supervised direction) on
family A, fix its polarity, and evaluate on the held-out family B, and B$¥to$A (Fig.‾¥ref{fig:cross}):

¥begin{center}
¥begin{tabular}{lcc}¥toprule
 & within-family (selection-inflated) & ¥textbf{cross-family (held out)} ¥¥¥midrule
best SAE feature & A $1.0$, B $1.0$ & ¥textbf{A$¥to$B $0.50$}, B$¥to$A $0.99$ ¥¥
supervised $d$ (on cells) & A $1.0$, B $0.99$ & A$¥to$B $0.76$, B$¥to$A $1.0$ ¥¥¥bottomrule
¥end{tabular}
¥end{center}

The A-selected feature ¥emph{does not transfer} (A$¥to$B $=0.50$, chance): it read a lexical shortcut present
only in family A (the filename, ``delete'', ``authorize''). A feature ¥emph{does} transfer B$¥to$A ($0.99$), and
a pooled feature separates granted from felt across both families (AUROC $0.998$, $p<0.001$) --- which, taken
alone, looks like a paraphrase-invariant semantic signal. ¥textbf{Naming settles it}: those ``invariant''
features (¥#34355, ¥#10017) max-activate on tool-call ¥emph{format boilerplate}
(¥texttt{function=example¥_function¥_name><parameter=example¥_parameter¥_}) and the ¥texttt{<function=}
scaffold --- present in every commit, with context-modulated magnitude that happens to correlate with the
condition. (Our pre-registered worry that the invariant signal was an em-dash punctuation confound was
¥emph{wrong}; it is scaffold, not punctuation --- the conclusion held for a different reason.) The separation is
surface, not a represented grant.

¥begin{figure}[t]¥centering
¥includegraphics[width=0.82¥textwidth]{figures_co/fig2_crosslexical.pdf}
¥caption{Cross-lexical-family generalization (held-out, selection-free). The lexical shortcut collapses to
chance (A$¥to$B $0.50$); the feature that transfers is tool-call scaffold (Fig.‾¥ref{fig:naming}), not
authorization.}
¥label{fig:cross}
¥end{figure}

¥section{Discussion}
¥label{sec:disc}
¥paragraph{Form, not granted.} Across three independent confound controls --- conversation structure (the
flip), user-turn lexicon (the cross-family collapse), and scaffold activation (the naming) --- the separator of
granted from felt commits is always ¥emph{surface form}. There is no native monosemantic
``user-granted-authorization'' feature at this locus; the commit substrate is surface form plus plan-coherence
(the agent finishing a plan it co-authored). This deepens ``felt, not granted''‾¥cite{feltnotgranted}: the
internal signal does not even track ¥emph{felt} authorization ¥emph{semantically} --- it tracks ¥emph{form}.
¥paragraph{Why internal monitors fail.} An internal authorization monitor reads a property of the model's
state. If the safety-relevant abstraction (``did the trusted user grant this specific irreversible action'')
is ¥emph{not} a feature the model computes, no probe or steering vector can read or move it; the apparent
``authorization direction'' was reading the form that correlates with it in a given prompt distribution, which
is why it generalizes poorly and inverts under a format change. The implication for agent safety is not
``calibrate the probe better'' but ``the grant must be checked ¥emph{outside} the model'', against the explicit
request --- consistent with the external-check result of‾¥cite{feltnotgranted}.
¥paragraph{Methodological note.} The automated best-feature flag said ¥emph{native feature found} in every run;
the pre-registered naming and the cross-lexical control overturned it each time. Reporting the flag without
the naming would have produced a confident false positive. This is the discipline, not a footnote.

¥section{Limitations}
This is one model (Qwen3.6-27B), one locus (layer 59), one domain (file deletion), with synthetic
structure-matched scenarios and an SAE that explains $¥approx 66¥%$ of residual variance. ``No native
authorization feature'' is, strictly, ¥emph{no monosemantic separating feature in this SAE's learned
dictionary at this locus}; a stronger SAE, a different layer, or a non-linear / multi-feature code could
surface a representation this analysis misses --- absence of a separating feature is weaker than evidence of
absence of any representation. The robust, SAE-independent result is the ¥emph{flip} (¥S¥ref{sec:flip}); the
SAE autopsy explains ¥emph{why} the direction is structural. The synthetic granted/felt distinction is itself
a design choice; interchange interventions on real long-horizon trajectories are the natural next test. We do
not re-test the steering-control of‾¥cite{authdir}.

¥section{Conclusion}
Pointing a sparse autoencoder at an agent's commit point, instead of a supervised difference-of-means
direction, overturns the clean reading of the ``authorization direction'': it is confounded by conversation
structure (it inverts, $0.838¥to0.08$), and the features that separate granted from felt commits are
surface-form --- tool-call scaffold, schema, file-listing --- with no native monosemantic representation of
user-granted authorization. The agent commits on form and plan-coherence. Internal authorization monitors
read form, not grant; the grant must be checked outside the model.

¥paragraph{Reproducibility.} Runners (¥texttt{agentguard¥_sae¥_¥{autopsy,autopsy¥_matched,counterfactual,name¥}.py}),
five per-run ledgers, figures, and an evaluation that recomputes every number are released at
¥url{https://github.com/OpenInterpretability/agentguard}; ledgers also at the Hugging Face dataset
¥texttt{caiovicentino1/swebench-phase6-verdict-circuit}. The decisive counterfactual run reproduced
bit-for-bit across two independent GPU sessions.

¥begin{thebibliography}{99}
¥bibitem{authdir} C.‾Vicentino. The Authorization Direction. OpenInterpretability, 2026. DOI:10.5281/zenodo.20683623.
¥bibitem{feltnotgranted} C.‾Vicentino. Felt, Not Granted. OpenInterpretability, 2026. DOI:10.5281/zenodo.20685264.
¥bibitem{refusaldir} A.‾Arditi, O.‾Obeso, A.‾Syed, D.‾Paleka, N.‾Panickssery, W.‾Gurnee, N.‾Nanda. Refusal in Language Models Is Mediated by a Single Direction. NeurIPS 2024. arXiv:2406.11717.
¥bibitem{monosemantic} T.‾Bricken, A.‾Templeton, J.‾Batson, et al. Towards Monosemanticity: Decomposing Language Models With Dictionary Learning. Transformer Circuits, 2023.
¥bibitem{saefind} H.‾Cunningham, A.‾Ewart, L.‾Riggs, R.‾Huben, L.‾Sharkey. Sparse Autoencoders Find Highly Interpretable Features in Language Models. ICLR 2024. arXiv:2309.08600.
¥bibitem{topk} L.‾Gao, T.‾Dupr¥'e la Tour, H.‾Tillman, G.‾Goh, R.‾Troll, A.‾Radford, I.‾Sutskever, J.‾Leike, J.‾Wu. Scaling and Evaluating Sparse Autoencoders. 2024. arXiv:2406.04093.
¥bibitem{geometry} S.‾Marks, M.‾Tegmark. The Geometry of Truth: Emergent Linear Structure in LLM Representations of True/False Datasets. 2023. arXiv:2310.06824.
¥bibitem{interpillusion} T.‾Bolukbasi, A.‾Pearce, A.‾Yuan, A.‾Coenen, E.‾Reif, F.‾Vi¥'egas, M.‾Wattenberg. An Interpretability Illusion for BERT. 2021. arXiv:2104.07143.
¥end{thebibliography}
¥end{document}
